---
title: "How effectively can LSTM networks predict impending software-system failures from multivariate telemetry"
subtitle: "This is the capstone project of IDC6910 class"
author: "Duc Nguyen, Cameron Ball, Luis Madrigal (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

Modern software systems generate large amounts of telemetry data, including
performance metrics, logs, and other operational signals. This information may
contain early warning signs of system failures; however, detecting those signs
can be challenging, especially in complex systems. Improving failure prediction
can help organizations respond earlier, reduce system downtime and associated
costs, and improve overall system reliability.

This project will use Microsoft's OpenRCA dataset [@xu2025openrca] to
investigate whether Long Short-Term Memory (LSTM) neural networks can predict
impending software-system failures from system telemetry. Because LSTMs are
designed to learn patterns over time, they may be useful for detecting changes
in telemetry data that occur before a system failure. We will evaluate how
accurately failures can be predicted at different time horizons and whether
combining multiple telemetry sources improves prediction performance compared
with using a single source. Overall, this project applies machine learning to a
practical software reliability problem by testing the effectiveness of telemetry
data for early failure detection.

Previous research is supportive of using LSTM models in the analysis of sequential data.
[@Gonzalez2018] showed that LSTM networks can be used in modeling nonlinear systems,
while [@ErgenKozat2020] used LSTM-based methods for anomaly detection in data sequences
with variable length. Other studies applied LSTMs to telemetry data obtained from computer
systems. [@MudgalWouhaybi2024] used a combined LSTM encoder in anomaly detection methods;
these include Isolation Forest, used to identify PC system failures from CPU, memory, disk,
temperature, and system data. This study obtained results that suggested LSTM-based telemetry
analysis can provide useful results for system-failure detection.

System log research also shows how important sequential modeling can be in failure-related
tasks. [@Chen2022] used two LSTM networks in the analysis of events before and after a log
entry; this improved anomaly detection in long log sequences. In addition, [@Hadadi2024]
researched deep learning for log-based failure prediction and found that model performance is
dependent on how log data are represented and transformed into sequences. The OpenRCA dataset
will be particularly useful in this project as it contains real failure case data from enterprise
systems, with multiple sources of telemetry, such as logs, metrics, and traces [@Xu2025].
While the prior work mentioned mainly focused on anomaly detection or cause analysis, this
project will be more focused on predicting failures before they occur and testing the efficacy
of multiple telemetry sources in improving early failure prediction performance.

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
